Adaptive broadcast tree for multicores

### 2013-05-21 paper.tex
Rewrote text about synchrony and added a list of MPI communication primitives

Atomic broadcast is a building-block for many higher-level distributed
algorithm such as state-machine replication. Such distributed
algorithms are increasingly used to overcome scalability challenges on
multicore machines. Examples are databases~\cite{Salomie2011,
  Wiesmann2000} and operating systems~\cite{fos:osr09, tornado:osdi99,
  barrelfish:sosp09}.

@@

% New ideas
Instead of manually tuning algorithms to characteristics of individual
machines, we automatically configure our atomic broadcast
implementation based on a machine model. We represent this model as an
annotated graph. One question we are looking at is which information
should actually be part of this machine model.

% How will you go about it?
We will base our atomic broadcast on \emph{overlay networks}, which we
automatically configure. We execute our configuration stage on a model
representing the concrete machine the algorithm is intended to run
on. %
Our idea is to then explore different approaches for finding overlay
networks based on the network model graph. Approaches we are currently
investigating are: minimum spanning tree algorithms, binary trees,
clustering techniques as well as treating the problem
hierarchically. Assuming the overlay network, we then need to decide
on every node which message to send first. This is a \emph{scheduling}
problem.

% How do we show it works
We auto-configure a wide range of different multicore machines and
compare the performance of our implementation with hand-tuned
implementations on all of these machines.
% Hypothesis
Our hypothesis is that automatic tuning of atomic broadcast (and other
distributed algorithms) to machine characteristics is worth
doing. Obviously, we cannot achieve performance of hand-tuned
implementations, but we hope to get close enough to argue that loosing
a bit of performance to avoid tedious manual reconfiguration is worth
if for many applications (while others, where performance
\emph{really} matters, might still want to manually configure their
stuff).

@@

@@

\begin{description}
\item[NUMA affinity] A list of cores on the same NUMA node.
\item[synchrony] Parts of a multicore machine are synchronous, i.e.\
  cores execute in lock-steps, where every node can send a message to
  every other node in one such step. This implies that there is a
  global time and message delivery time is bound. Both can be assumed
  on today's multicores. The global time comes from the fact that
  cores are driven by the same quartz or, if not, they are
  synchronized such that the jitter is less than a cycle, which leads
  to cores seeing the same time. %
  For these reasons, synchrony is not part of our current machine
  model. Within today's multicores, synchrony can be
  assumed. \stefan{Can it?}
\item[failure] Can nodes fail? If so, how? (Byzantine, crash). We do
  not consider failures at the moment, but believe that we will
  inevitably have to do so for future multicores, where it is unlikely
  that hardware can hide failures from software.
\end{description}

@@

\begin{itemize}
\item Arguing that group communication/atomic broadcast is important
  and showing that we can do it well might be enough (appeal to atomic
  broadcasts as the foundation for other distributed algorithms, and
  the MPI primitives, that can/need to be mapped to
  group-communication (have a list of them)
\item Need higher-level applications, such as:
  \begin{itemize}
  \item capability system? \stefan{Ask Simon how expensive revocation
      currently is}
  \item a database with replication and consistency maintenance
    \begin{itemize}
    \item SharedDB? (buffers between operators, but no synchronization
      except for access to buffers, which are multiple writers, one
      reader)
    \item Crescando (some kind of state machine replication)
    \end{itemize}
  \end{itemize}
\item MPI collectives (as in~\cite{Tu2008})
\item \emph{Load balancing in a tree} We could build several rings,
  that do not always include all coordinators and work with read- and
  write-sets. We can select these rings such that they nicely
  integrate with the physical topology. They can also have different
  sizes (e.g.\ a smaller one for read operations and larger ones for
  writes).
\item \emph{Aggregation for convergecast} %
  Convergecast is a way of collecting information from nodes (i.e.\ it
  is kind of the reverse of a broadcast). On the return path,
  information can be aggregated. One example is batching, where several
  pieces of information are grouped in the same message to reduce packet
  processing overhead. Another example is aggregation, i.e.\
  pre-processing of data. One example is pre-calculating the average
  value of child messages if the sink node is not interested in
  individual values. %
  Similarly to what has been done in wireless sensor networks (where it
  also matters to reduce the number of messages, but for other reasons:
  power consumption), we can do aggregation in nodes. In difference to
  traditional distributed systems, this works, because it is easy to
  deploy custom software on every node in the network. Classical
  distributed systems do not typically allow this. Furthermore, reducing
  the number of messages at the price of higher complexity does not make
  sense in classical systems. %
  Examples for aggregation: number of nodes agreeing to something, find
  capabilities (concatenate core ids). 
\end{itemize}

>>>>

Distributed algorithms are increasingly used to overcome scalability
challenges on multicore machines. Examples are
databases~\cite{Salomie2011, Wiesmann2000} and operating
systems~\cite{fos:osr09, tornado:osdi99, barrelfish:sosp09}.

@@

% New ideas
Instead of manually tuning algorithms to characteristics of individual
machines, we automatically configure distributed algorithm
implementations based on a machine model. We represent this model as an
annotated graph. One question we are looking at is which information
should actually be part of this machine model.

% How will you go about it?
We base the configuration of algorithms on \emph{overlay networks}, which we
automatically configure. We execute our configuration stage on a model
representing the concrete machine the algorithm is intended to run
on. %
Our idea is to then explore different approaches for finding overlay
networks based on the network model graph. Approaches we are currently
investigating are: minimum spanning tree algorithms, binary trees,
clustering techniques as well as treating the problem
hierarchically. Assuming the overlay network, we then need to decide
on every node which message to send first. This is a \emph{scheduling}
problem.

% How do we show it works
We auto-configure a wide range of different multicore machines and
compare the performance of our implementation with hand-tuned
implementations on all of these machines.
% Hypothesis
Our hypothesis is that automatic tuning of distributed algorithms 
to machine characteristics is worth
doing. Obviously, we cannot achieve performance of hand-tuned
implementations, but we hope to get close enough to argue that loosing
a bit of performance to avoid tedious manual reconfiguration is worth
if for many applications (while others, where performance
\emph{really} matters, might still want to manually configure their
stuff).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Distributed algorithms}
\label{sec:distalgo}

In order to evaluate our ideas, we pick two concrete distributed
algorithms and implement them for a multicore machine:

\paragraph{Atomic broadcast} \stefan{Explain the guarantees provided
  by atomic broadcasts}. Atomic broadcasts are frequently used to
implement higher-level distributed algorithms on top of it: examples
are replication such as state machine replication.

\paragraph{Barriers} \stefan{Explain}

@@

\begin{description}
\item[NUMA affinity] A list of cores on the same NUMA node.
\item[Synchrony] Synchronous systems are characterized by two
  attributes: First, cores execute their programs in lock-steps and
  second that message delivery time is bound.
  % 
  The existence of lock-steps implies the existence of a global time.
  This is often true on multicore machines since cores are driven by
  the same quartz or, if not, they are synchronized such that the
  jitter is less than a cycle, which leads to cores seeing the same
  time. %
  Bounded delivery time is harder to reason about. CPUs on multicores
  are typically time-multiplexed. On such systems, several programs
  execute on the same processing unit one at a time. A scheduler
  decides the order in which to run these applications. In very
  contented systems, timely receipt of messages cannot be guaranteed
  globally for all applications since many of them compete for a CPU
  core. Even if we could guarantee timely delivery of messages by use
  of a deterministic scheduler, worst-case propagation times would
  still be orders of magnitude higher than best-case propagation
  (which is essentially zero). \stefan{gang-scheduling?}
\item[Failure] Can nodes fail? If so, how? (Byzantine, crash). We do
  not consider failures at the moment, but believe that we will
  inevitably have to do so for future multicores, where it is unlikely
  that hardware can hide failures from software.
\end{description}

@@

\begin{itemize}
\item Think about synchronous implementations and see if they are
  applicable for some of our machines.
\item Arguing that group communication/atomic broadcast is important
  and showing that we can do it well might be enough (appeal to atomic
  broadcasts as the foundation for other distributed algorithms, and
  the MPI primitives, that can/need to be mapped to
  group-communication (have a list of them)
  \begin{itemize}
  \item MPI_Bcast: send a broadcast message to every node (including
    the sending one)
  \item MPI_Scatter
  \item MPI_Gather
  \item MPI_Reduce
  \end{itemize}
  These primitives are synchronous. They return only once every other
  node received the message. 
\item Need higher-level applications, such as:
  \begin{itemize}
  \item capability system? \stefan{Ask Simon how expensive revocation
      currently is}
  \item a database with replication and consistency maintenance
    \begin{itemize}
    \item SharedDB? (buffers between operators, but no synchronization
      except for access to buffers, which are multiple writers, one
      reader)
    \item Crescando (some kind of state machine replication)
    \end{itemize}
  \end{itemize}
\item MPI collectives (as in~\cite{Tu2008})
\item \emph{Load balancing in a tree} We could build several rings,
  that do not always include all coordinators and work with read- and
  write-sets. We can select these rings such that they nicely
  integrate with the physical topology. They can also have different
  sizes (e.g.\ a smaller one for read operations and larger ones for
  writes).
\item \emph{Aggregation for convergecast} %
  Convergecast is a way of collecting information from nodes (i.e.\ it
  is kind of the reverse of a broadcast). On the return path,
  information can be aggregated. One example is batching, where several
  pieces of information are grouped in the same message to reduce packet
  processing overhead. Another example is aggregation, i.e.\
  pre-processing of data. One example is pre-calculating the average
  value of child messages if the sink node is not interested in
  individual values. %
  Similarly to what has been done in wireless sensor networks (where it
  also matters to reduce the number of messages, but for other reasons:
  power consumption), we can do aggregation in nodes. In difference to
  traditional distributed systems, this works, because it is easy to
  deploy custom software on every node in the network. Classical
  distributed systems do not typically allow this. Furthermore, reducing
  the number of messages at the price of higher complexity does not make
  sense in classical systems. %
  Examples for aggregation: number of nodes agreeing to something, find
  capabilities (concatenate core ids). 
\end{itemize}
